---
title: "C173 Project"
author: "Stephen Fujiwara"
output:
  pdf_document: default
date: "2025-03-01"
---

## EDA (Exploratory Data Analysis)

### Variable Description

```{r}
# read data
data <- read.csv("data.csv")
head(data)
```

The dataset is a dataset from the Los Angeles county database that contains census data from each of the zip codes that *fall at least partially* within Los Angeles county boundaries, for the year 2010.

Each observation contains the zip code, the coordinates corresponding to the centroid of the zip code's area, as well as 6 variables.

These 6 variables are:

1.  population - The total number of people living in a given zip code.

2.  median_age - The median age of the individuals living in a given zip code.

3.  percent_male - The percentage of the population that is male for a given a zip code.

4.  percent_female - The percentage of the population that is female for a given zip code.

5.  total_households - The total number of households located within a given zip code

6.  avg_household_size - The average number of individuals living in a household located within a given zip code.

### Descriptive Statistics

```{r}
# descriptive statistics
variables <- c("population", "median_age", "percent_male", "percent_female", "total_households", "avg_household_size")

summary(data[, variables])
```

### Histogram of Each Variable

```{r}
# histograms
par(mfrow = c(2, length(variables) / 2))
for (variable in variables) {
  hist(data[, variable], main = paste(variable, "Histogram"), xlab = variable)
}
```

### ECDF of Each Variable

```{r}
# ECDFs
par(mfrow = c(2, length(variables) / 2))
for (variable in variables) {
  plot(ecdf(data[, variable]), main = paste(variable, "ECDF"))
}
```

### Scatterplot of Each Variable (Individual)

```{r}
# scatterplots
par(mfrow = c(2, length(variables) / 2))
for (variable in variables) {
  plot(data[, variable], main = paste(variable, "Scatterplot"), ylab = variable)
}
```

### Scatterplot of Each Variable (Pair-wise)

```{r}
# scatterplots (pair-wise)
pairs(data[, variables], main = "Pair-wise Scatter Plots")
```

### Bubble Plot of Each Variable

```{r}
library(sp)
```

```{r}
# bubble plots of each variable

coordinates(data) <- ~ longitude + latitude

bubble(data, "population", maxsize = 1, main = "Population", xlab = "Longitude", ylab = "Latitude")

bubble(data, "median_age", maxsize = 1, main = "Median Age", xlab = "Longitude", ylab = "Latitude")

bubble(data, "percent_male", maxsize = 1, main = "Percent Male", xlab = "Longitude", ylab = "Latitude")

bubble(data, "percent_female", maxsize = 1, main = "Percent Female", xlab = "Longitude", ylab = "Latitude")

bubble(data, "total_households", maxsize = 1, main = "Total Households", xlab = "Longitude", ylab = "Latitude")

bubble(data, "avg_household_size", maxsize = 1, main = "Average Household Size", xlab = "Longitude", ylab = "Latitude")

# revert to regular dataframe
data <- as.data.frame(data)
```

### H-scatterplot of Each Variable

```{r}
library(gstat)
```

```{r}
# h-scatter plots of each variable

# convert to spatial dataframe object
coordinates(data) <-  ~ longitude + latitude

# compute max distance between coordinates of the dataset
max_distance <- max(dist(coordinates(data)))

# compute sequence of h-values to plot with
h <- seq(0, max_distance, length.out = 10)

hscat(population ~ 1, data, h)

hscat(median_age ~ 1, data, h)

hscat(percent_male ~ 1, data, h)

hscat(percent_female ~ 1, data, h)

hscat(total_households ~ 1, data, h)

hscat(avg_household_size ~ 1, data, h)

# convert back to regular dataframe
data <- as.data.frame(data)
```

## Variogram Modeling

### Fitting a Model to the Empirical Semivariogram (Classical)

#### Computing the Empirical Semivariogram (Classical)

```{r}
library(geoR)
```

```{r}
# create dataframe of just coordinates and target variable
target_variable <- "population"
target_data <- data[, c("longitude", "latitude", target_variable)]
head(target_data)
```

```{r}
# set target variable and create geodata object
geo_data <- as.geodata(target_data)
```

```{r}
# compute empirical semivariogram (classical)
empirical_semivariogram_c <- variog(geo_data, max.dist = 1.7)
plot(empirical_semivariogram_c)
```

#### Fitting a Spherical Model

```{r}
# fitting a model to the empirical semivariogram (classical) - spherical
plot(empirical_semivariogram_c)

cov_model <- "sph"
cov_pars <- c(3*10^8, 1.4)
nugget <- 4*10^8

# fitting by eye
lines.variomodel(cov.model = cov_model, cov.pars = cov_pars, nugget = nugget)

# fitting using equal weights
equal_weights_model <- variofit(empirical_semivariogram_c, cov.model = cov_model, ini.cov.pars = cov_pars, weights = "equal", fix.nugget = FALSE, nugget = nugget)

lines(equal_weights_model, lty = 1, col = "red")

# fitting using npairs weights
npairs_weights_model <- variofit(empirical_semivariogram_c, cov.model = cov_model, ini.cov.pars = cov_pars, weights = "npairs", fix.nugget = FALSE, nugget = nugget)

lines(npairs_weights_model, lty = 1, col = "blue")

# fitting using cressie's weights
cressies_weights_model <- variofit(empirical_semivariogram_c, cov.model = cov_model, ini.cov.pars = cov_pars, weights = "cressie", fix.nugget = FALSE, nugget = nugget)

lines(cressies_weights_model, lty = 1, col = "green")

# legend
legend(
  "bottomright",
  legend = c("By Eye", "Equal Weights", "Npairs", "Cressie"),
  col = c("black", "red", "blue", "green"),
  lty = 1
)
```

```{r}
# fitting a model using MLE - spherical
# although the empirical semivariogram is not used for fitting, we would like to observe it's fit to the empirical semivariogram (classical)
plot(empirical_semivariogram_c)

cov_model <- "sph"
cov_pars <- c(3*10^8, 1.4)
nugget <- 4*10^8

# fitting using MML
mml_model <- likfit(geo_data, cov.model = cov_model, ini.cov.pars = cov_pars,
       fix.nugget = FALSE, nugget = nugget)

lines(mml_model, lty = 1, col="orange")

# fitting using REML
reml_model <- likfit(geo_data, cov.model = cov_model, ini.cov.pars = cov_pars,
       fix.nugget = FALSE, nugget = nugget, lik.method = "RML")
lines(reml_model, lty = 1, col = "purple")

# legend
legend(
  "bottomright",
  legend = c("MML", "REML"),
  col = c("orange", "purple"),
  lty = 1
)
```

#### Fitting an Exponential Model

```{r}
# fitting a model to the empirical semivariogram (classical) - exponential
plot(empirical_semivariogram_c)

cov_model <- "exp"
cov_pars <- c(3*10^8, 1.4)
nugget <- 4*10^8

# fitting by eye
lines.variomodel(cov.model = cov_model, cov.pars = cov_pars, nugget = nugget)

# fitting using equal weights
equal_weights_model <- variofit(empirical_semivariogram_c, cov.model = cov_model, ini.cov.pars = cov_pars, weights = "equal", fix.nugget = FALSE, nugget = nugget)

lines(equal_weights_model, lty = 1, col = "red")

# fitting using npairs weights
npairs_weights_model <- variofit(empirical_semivariogram_c, cov.model = cov_model, ini.cov.pars = cov_pars, weights = "npairs", fix.nugget = FALSE, nugget = nugget)

lines(npairs_weights_model, lty = 1, col = "blue")

# fitting using cressie's weights
cressies_weights_model <- variofit(empirical_semivariogram_c, cov.model = cov_model, ini.cov.pars = cov_pars, weights = "cressie", fix.nugget = FALSE, nugget = nugget)

lines(cressies_weights_model, lty = 1, col = "green")

# legend
legend(
  "bottomright",
  legend = c("By Eye", "Equal Weights", "Npairs", "Cressie"),
  col = c("black", "red", "blue", "green"),
  lty = 1
)
```

```{r}
# fitting a model using MLE - exponential
# although the empirical semivariogram is not used for fitting, we would like to observe it's fit to the empirical semivariogram (classical)
plot(empirical_semivariogram_c)

cov_model <- "exp"
cov_pars <- c(3*10^8, 1.4)
nugget <- 4*10^8

# fitting using MML
mml_model <- likfit(geo_data, cov.model = cov_model, ini.cov.pars = cov_pars,
       fix.nugget = FALSE, nugget = nugget)

lines(mml_model, lty = 1, col="orange")

# fitting using REML
reml_model <- likfit(geo_data, cov.model = cov_model, ini.cov.pars = cov_pars,
       fix.nugget = FALSE, nugget = nugget, lik.method = "RML")
lines(reml_model, lty = 1, col = "purple")

# legend
legend(
  "bottomright",
  legend = c("MML", "REML"),
  col = c("orange", "purple"),
  lty = 1
)
```

#### Fitting a Gaussian Model

```{r}
# fitting a model to the empirical semivariogram (classical) - gaussian
plot(empirical_semivariogram_c)

cov_model <- "gauss"
cov_pars <- c(3*10^8, 1.4)
nugget <- 4*10^8

# fitting by eye
lines.variomodel(cov.model = cov_model, cov.pars = cov_pars, nugget = nugget)

# fitting using equal weights
equal_weights_model <- variofit(empirical_semivariogram_c, cov.model = cov_model, ini.cov.pars = cov_pars, weights = "equal", fix.nugget = FALSE, nugget = nugget)

lines(equal_weights_model, lty = 1, col = "red")

# fitting using npairs weights
npairs_weights_model <- variofit(empirical_semivariogram_c, cov.model = cov_model, ini.cov.pars = cov_pars, weights = "npairs", fix.nugget = FALSE, nugget = nugget)

lines(npairs_weights_model, lty = 1, col = "blue")

# fitting using cressie's weights
cressies_weights_model <- variofit(empirical_semivariogram_c, cov.model = cov_model, ini.cov.pars = cov_pars, weights = "cressie", fix.nugget = FALSE, nugget = nugget)

lines(cressies_weights_model, lty = 1, col = "green")

# legend
legend(
  "bottomright",
  legend = c("By Eye", "Equal Weights", "Npairs", "Cressie"),
  col = c("black", "red", "blue", "green"),
  lty = 1
)
```

```{r}
# fitting a model using MLE - gaussian
# although the empirical semivariogram is not used for fitting, we would like to observe it's fit to the empirical semivariogram (classical)
plot(empirical_semivariogram_c)

cov_model <- "gauss"
cov_pars <- c(3*10^8, 1.4)
nugget <- 4*10^8

# fitting using MML
mml_model <- likfit(geo_data, cov.model = cov_model, ini.cov.pars = cov_pars,
       fix.nugget = FALSE, nugget = nugget)

lines(mml_model, lty = 1, col="orange")

# fitting using REML
reml_model <- likfit(geo_data, cov.model = cov_model, ini.cov.pars = cov_pars,
       fix.nugget = FALSE, nugget = nugget, lik.method = "RML")
lines(reml_model, lty = 1, col = "purple")

# legend
legend(
  "bottomright",
  legend = c("MML", "REML"),
  col = c("orange", "purple"),
  lty = 1
)
```

### Fitting a Model to the Empirical Semivariogram (Robust)

#### Computing the Empirical Semivariogram (Robust)

```{r}
# compute empirical semivariogram (robust)
empirical_semivariogram_r <- variog(geo_data, estimator.type = "modulus" , max.dist = 1.7)
plot(empirical_semivariogram_r)
```

#### Fitting a Spherical Model

```{r}
# fitting a model to the empirical semivariogram (robust) - spherical
plot(empirical_semivariogram_r)

cov_model <- "sph"
cov_pars <- c(4*10^8, 1.4)
nugget <- 4*10^8

# fitting by eye
lines.variomodel(cov.model = cov_model, cov.pars = cov_pars, nugget = nugget)

# fitting using equal weights
equal_weights_model <- variofit(empirical_semivariogram_r, cov.model = cov_model, ini.cov.pars = cov_pars, weights = "equal", fix.nugget = FALSE, nugget = nugget)

lines(equal_weights_model, lty = 1, col = "red")

# fitting using npairs weights
npairs_weights_model <- variofit(empirical_semivariogram_r, cov.model = cov_model, ini.cov.pars = cov_pars, weights = "npairs", fix.nugget = FALSE, nugget = nugget)

lines(npairs_weights_model, lty = 1, col = "blue")

# fitting using cressie's weights
cressies_weights_model <- variofit(empirical_semivariogram_r, cov.model = cov_model, ini.cov.pars = cov_pars, weights = "cressie", fix.nugget = FALSE, nugget = nugget)

lines(cressies_weights_model, lty = 1, col = "green")

# legend
legend(
  "bottomright",
  legend = c("By Eye", "Equal Weights", "Npairs", "Cressie"),
  col = c("black", "red", "blue", "green"),
  lty = 1
)
```

```{r}
# fitting a model using MLE - spherical
# although the empirical semivariogram is not used for fitting, we would like to observe it's fit to the empirical semivariogram (robust)

plot(empirical_semivariogram_r)

cov_model <- "sph"
cov_pars <- c(4*10^8, 1.4)
nugget <- 4*10^8

# fitting using MML
mml_model <- likfit(geo_data, cov.model = cov_model, ini.cov.pars = cov_pars,
       fix.nugget = FALSE, nugget = nugget)

lines(mml_model, lty = 1, col="orange")

# fitting using REML
reml_model <- likfit(geo_data, cov.model = cov_model, ini.cov.pars = cov_pars,
       fix.nugget = FALSE, nugget = nugget, lik.method = "RML")
lines(reml_model, lty = 1, col = "purple")

# legend
legend(
  "bottomright",
  legend = c("MML", "REML"),
  col = c("orange", "purple"),
  lty = 1
)
```

#### Fitting an Exponential Model

```{r}
# fitting a model to the empirical semivariogram (robust) - exponential
plot(empirical_semivariogram_r)

cov_model <- "exp"
cov_pars <- c(4*10^8, 1.4)
nugget <- 4*10^8

# fitting by eye
lines.variomodel(cov.model = cov_model, cov.pars = cov_pars, nugget = nugget)

# fitting using equal weights
equal_weights_model <- variofit(empirical_semivariogram_r, cov.model = cov_model, ini.cov.pars = cov_pars, weights = "equal", fix.nugget = FALSE, nugget = nugget)

lines(equal_weights_model, lty = 1, col = "red")

# fitting using npairs weights
npairs_weights_model <- variofit(empirical_semivariogram_r, cov.model = cov_model, ini.cov.pars = cov_pars, weights = "npairs", fix.nugget = FALSE, nugget = nugget)

lines(npairs_weights_model, lty = 1, col = "blue")

# fitting using cressie's weights
cressies_weights_model <- variofit(empirical_semivariogram_r, cov.model = cov_model, ini.cov.pars = cov_pars, weights = "cressie", fix.nugget = FALSE, nugget = nugget)

lines(cressies_weights_model, lty = 1, col = "green")

# legend
legend(
  "bottomright",
  legend = c("By Eye", "Equal Weights", "Npairs", "Cressie"),
  col = c("black", "red", "blue", "green"),
  lty = 1
)
```

```{r}
# fitting a model using MLE - exponential
# although the empirical semivariogram is not used for fitting, we would like to observe it's fit to the empirical semivariogram (robust)

plot(empirical_semivariogram_r)

cov_model <- "exp"
cov_pars <- c(4*10^8, 1.4)
nugget <- 4*10^8

# fitting using MML
mml_model <- likfit(geo_data, cov.model = cov_model, ini.cov.pars = cov_pars,
       fix.nugget = FALSE, nugget = nugget)

lines(mml_model, lty = 1, col="orange")

# fitting using REML
reml_model <- likfit(geo_data, cov.model = cov_model, ini.cov.pars = cov_pars,
       fix.nugget = FALSE, nugget = nugget, lik.method = "RML")
lines(reml_model, lty = 1, col = "purple")

# legend
legend(
  "bottomright",
  legend = c("MML", "REML"),
  col = c("orange", "purple"),
  lty = 1
)
```

#### Fitting a Gaussian Model

```{r}
# fitting a model to the empirical semivariogram (robust) - gaussian
plot(empirical_semivariogram_r)

cov_model <- "gaussian"
cov_pars <- c(4*10^8, 1.4)
nugget <- 4*10^8

# fitting by eye
lines.variomodel(cov.model = cov_model, cov.pars = cov_pars, nugget = nugget)

# fitting using equal weights
equal_weights_model <- variofit(empirical_semivariogram_r, cov.model = cov_model, ini.cov.pars = cov_pars, weights = "equal", fix.nugget = FALSE, nugget = nugget)

lines(equal_weights_model, lty = 1, col = "red")

# fitting using npairs weights
npairs_weights_model <- variofit(empirical_semivariogram_r, cov.model = cov_model, ini.cov.pars = cov_pars, weights = "npairs", fix.nugget = FALSE, nugget = nugget)

lines(npairs_weights_model, lty = 1, col = "blue")

# fitting using cressie's weights
cressies_weights_model <- variofit(empirical_semivariogram_r, cov.model = cov_model, ini.cov.pars = cov_pars, weights = "cressie", fix.nugget = FALSE, nugget = nugget)

lines(cressies_weights_model, lty = 1, col = "green")

# legend
legend(
  "bottomright",
  legend = c("By Eye", "Equal Weights", "Npairs", "Cressie"),
  col = c("black", "red", "blue", "green"),
  lty = 1
)
```

```{r}
# fitting a model using MLE - gaussian
# although the empirical semivariogram is not used for fitting, we would like to observe it's fit to the empirical semivariogram (robust)

plot(empirical_semivariogram_r)

cov_model <- "gauss"
cov_pars <- c(4*10^8, 1.4)
nugget <- 4*10^8

# fitting using MML
mml_model <- likfit(geo_data, cov.model = cov_model, ini.cov.pars = cov_pars,
       fix.nugget = FALSE, nugget = nugget)

lines(mml_model, lty = 1, col="orange")

# fitting using REML
reml_model <- likfit(geo_data, cov.model = cov_model, ini.cov.pars = cov_pars,
       fix.nugget = FALSE, nugget = nugget, lik.method = "RML")
lines(reml_model, lty = 1, col = "purple")

# legend
legend(
  "bottomright",
  legend = c("MML", "REML"),
  col = c("orange", "purple"),
  lty = 1
)
```

## Spatial Prediction

### Predictions on a dense grid using the inverse distance interpolation method

```{r}
# create grid
x_min <- min(target_data$longitude)
x_max <- max(target_data$longitude)
y_min <- min(target_data$latitude)
y_max <- max(target_data$latitude)

x_values <- seq(x_min, x_max, length.out = 100)
y_values <- seq(y_min, y_max, length.out = 100)

xy_grid <- expand.grid(longitude = x_values, latitude = y_values)

coordinates(xy_grid) <- ~ longitude + latitude

# plot grid
plot(xy_grid)

# plot locations of observed data on the grid
points(target_data)
```

```{r}
# predict data points using grid as input
idw_pred <- idw(formula = population ~ 1, locations = ~ longitude + latitude, target_data, xy_grid)$var1.pred

# convert vector to matrix
idw_pred_matrix <- matrix(idw_pred, length(x_values), length(y_values))

# plot predicted values
image(x_values, y_values, idw_pred_matrix, xlab = "East to West", ylab = "South to North")

# plot locations of observed data on the grid
points(target_data)
```

### Predictions using different types of kriging

```{r}
# variogram model to use for kriging
model <- equal_weights_model
```

#### Simple Kriging

```{r}
# simple kriging
simple_krig <- krige.conv(geo_data, locations = as.data.frame(xy_grid), krige = krige.control(type.krige = "sk", obj.model = model))

head(simple_krig$predict)
```

#### Ordinary Kriging

```{r}
# ordinary kriging
ordinary_krig <- krige.conv(geo_data, locations = as.data.frame(xy_grid), krige = krige.control(type.krige = "ok", obj.model = model))

head(ordinary_krig$predict)
```

#### Universal Kriging

```{r}
# fit a variogram model with the assumption of trend

# convert to gstat object
g_stat <- gstat(formula = population ~ longitude + latitude, locations = ~ longitude + latitude, data = target_data)

# compute empirical semivariogram
empirical_semivariogram_trend <- variogram(g_stat, cutoff = 0.3)
plot(empirical_semivariogram_trend)

# fit a variogram model
trend_model <- fit.variogram(empirical_semivariogram_trend, vgm(5*10^8, "Sph", 0.2, 2*10^8))
plot(empirical_semivariogram_trend, trend_model)

# universal kriging
univ_krig <- krige(formula = population ~ longitude + latitude, locations = ~ longitude + latitude, model = trend_model, data = target_data, newdata = as.data.frame(xy_grid))

head(univ_krig$var1.pred)
```

#### Co-kriging

```{r}
# compute correlation of population with each of the other variables
variable_data <- data[, variables]
correlations <- cor(variable_data, use = "pairwise.complete.obs")["population", ]
correlations
```

```{r}
# conduct co-kriging with population as the target variable and two other co-located variables: total_households and avg_household_size

# create gstat object including all three variables
g_stat <- gstat(id="population", formula = population ~ 1, locations = ~ longitude + latitude, data = data)
g_stat <- gstat(g_stat, id = "total_households", formula = total_households ~ longitude + latitude, locations = ~ longitude + latitude, data = data)

# compute and plot the semivariograms for each of the variables
empirical_cross_semivariogram <- variogram(g_stat)
plot(empirical_cross_semivariogram)

# fit a variogram model to the sample variogram of the target variable
g_stat_target <- gstat(id = "population", formula = population ~ 1, locations = ~ longitude + latitude, data = target_data)
empirical_semivariogram_target <- variogram(g_stat_target, cutoff = 0.3)
plot(empirical_semivariogram_target)
model_target <- fit.variogram(empirical_semivariogram_target, vgm(4*10^8, "Exp", 0.2, 0), fit.method = 2)
plot(empirical_semivariogram_target, model_target)

# fit a variogram model to the sample cross-variogram
model_cross <- fit.lmc(empirical_cross_semivariogram, g_stat, model = model_target)
plot(empirical_cross_semivariogram, model_cross)

# co-kriging
co_krig <- predict(model_cross, xy_grid)
head(co_krig$population.pred)
```

### Raster Maps and add Contours, for each set of kriging predictions

```{r}
# simple kriging raster map
prediction_matrix <- matrix(simple_krig$predict, length(x_values), length(y_values))
image(x_values, y_values, prediction_matrix, xlab = "East to West", ylab = "South to North", main = paste("Raster Map of Simple Kriging Predictions w/ Contours"))
contour(x_values, y_values, prediction_matrix, color = "black", add = TRUE)

# ordinary kriging raster map
prediction_matrix <- matrix(ordinary_krig$predict, length(x_values), length(y_values))
image(x_values, y_values, prediction_matrix, xlab = "East to West", ylab = "South to North", main = paste("Raster Map of Ordinary Kriging Predictions w/ Contours"))
contour(x_values, y_values, prediction_matrix, color = "black", add = TRUE)

# universal kriging raster map
prediction_matrix <- matrix(univ_krig$var1.pred, length(x_values), length(y_values))
image(x_values, y_values, prediction_matrix, xlab = "East to West", ylab = "South to North", main = paste("Raster Map of Universal Kriging Predictions w/ Contours"))
contour(x_values, y_values, prediction_matrix, color = "black", add = TRUE)

# co-kriging raster map
prediction_matrix <- matrix(co_krig$population.pred, length(x_values), length(y_values))
image(x_values, y_values, prediction_matrix, xlab = "East to West", ylab = "South to North", main = paste("Raster Map of Co-Kriging Predictions w/ Contours"))
contour(x_values, y_values, prediction_matrix, color = "black", add = TRUE)
```
